---
title: "UFO Sigthings Near Area51- Visualization"
author: "Mrudula Nimmala, Sowmya Sree Kemsaram, Shreyas Shivaji Mali"
format:
    html:
     self-contained: true
---

## Loading the necessary libraries

```{r warning=FALSE, message=FALSE}

library(lubridate)
library(dplyr)
library(geosphere)
library(leaflet)
library(dbscan)
library(MASS)
library(ggplot2)
library(plotly)


```

## Loading the Data set and perform pre-processing of data

```{r}

# Loading the dataset
file_path <- "ufo-sightings-transformed.csv"
ufo_dataset <- read.csv(file_path)


# Overview of the dataset
head(ufo_dataset)
summary(ufo_dataset)

# Checking for missing values
colSums(ufo_dataset == "" | is.na(ufo_dataset))

# Removing rows with missing or blank values
ufo_dataset <- ufo_dataset[!(apply(ufo_dataset, 1, function(row) any(row == "" | is.na(row)))), ]

# Converting Date_time to datetime
ufo_dataset$Date_time <- ymd_hms(ufo_dataset$Date_time)

# Converting date_documented to date
ufo_dataset$date_documented <- mdy(ufo_dataset$date_documented)



# Converting columns to categorical variables (factors)
ufo_dataset$Season <- as.factor(ufo_dataset$Season)
ufo_dataset$Country_Code <- as.factor(ufo_dataset$Country_Code)
ufo_dataset$Country <- as.factor(ufo_dataset$Country)
ufo_dataset$Region <- as.factor(ufo_dataset$Region)
ufo_dataset$Locale <- as.factor(ufo_dataset$Locale)
ufo_dataset$UFO_shape <- as.factor(ufo_dataset$UFO_shape)

```

## Get Area-51 Coordinates. 

```{r}

# Coordinates for Area 51
area51_coords <- c(-115.808, 37.233)

# Function to calculate distance from Area 51
calculate_distance <- function(lon, lat) {
  # Converting to miles
  dist <- distGeo(c(lon, lat), area51_coords) / 1609.34  
  return(dist)
}

# Adding a distance column to the dataset
data <- ufo_dataset %>%
  mutate(Distance_to_Area51 = mapply(calculate_distance, longitude, latitude))

# Filtering for sightings within 50 miles of Area 51
area51_sightings <- data %>%
  filter(Distance_to_Area51 <= 50)

```

## Pointing the UFO sigthings near Area 51

```{r}
leaflet(data = area51_sightings) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    popup = ~paste(
      "Date and Time:", Date_time, "<br>",
      "Season:", Season, "<br>",
      "Region:", Region, "<br>",
      "Locale:", Locale
    ),
    radius = 5,
    color = "red",
    fillOpacity = 0.8
  ) %>%
  setView(lng = -115.808, lat = 37.233, zoom = 8)

```


```{r warning=FALSE}
library(plotly)
# Defining Area 51 coordinates
area51_coords <- data.frame(longitude = -115.808, latitude = 37.233)

# Adjusting the density plot and annotating Area 51
gg_density <- ggplot(area51_sightings, aes(x = longitude, y = latitude)) +
  geom_density2d() +
  stat_density2d(aes(fill = ..level..), geom = "polygon", alpha = 0.4) +
  geom_point(data = area51_coords, aes(x = longitude, y = latitude), 
             color = "red", size = 3) +  
  annotate("text", x = -115.808, y = 37.233, label = "Area 51", 
           color = "red", size = 5, hjust = 0, vjust = -1) + 
  labs(title = "Density of UFO Sightings Near Area 51",
       x = "Longitude", y = "Latitude") +
  xlim(-116, -115.5) +  
  ylim(36.5, 37.5) +    
  theme_minimal()

# Converting to an interactive plot
interactive_density <- ggplotly(gg_density)

# Displaying the plot
interactive_density



```

## Time-series patterns at AREA 51

### Extracting Date and Time From the Dataset

```{r}

area51_sightings$Year <- format(area51_sightings$Date_time, "%Y")
area51_sightings$Month <- format(area51_sightings$Date_time, "%m")
area51_sightings$Day <- format(area51_sightings$Date_time, "%d")
area51_sightings$Hour <- format(area51_sightings$Date_time, "%H")

```


### Analysing Sightings by Year

```{r}

# Grouping by year and count sightings
sightings_by_year <- area51_sightings %>%
  group_by(Year) %>%
  summarise(Sightings = n())

# Plotting the trend over years
ggplot(sightings_by_year, aes(x = as.numeric(Year), y = Sightings)) +
  geom_line(color = "blue", size = 1) +
  geom_point(size = 2) +
  labs(title = "UFO Sightings Near Area 51 Over the Years",
       x = "Year", y = "Number of Sightings") +
  theme_minimal()


```

### Analyze Sightings by Month

```{r}

# Grouping by month and count sightings
sightings_by_month <- area51_sightings %>%
  group_by(Month) %>%
  summarise(Sightings = n())

# Plotting the trend over months
ggplot(sightings_by_month, aes(x = Month, y = Sightings)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
  labs(title = "UFO Sightings Near Area 51 by Month",
       x = "Month", y = "Number of Sightings") +
  theme_minimal()


```

### Analyzing Sightings by Hour

```{r warning=FALSE}

# Grouping by hour and count sightings
sightings_by_hour <- area51_sightings %>%
  group_by(Hour) %>%
  summarise(Sightings = n())

# Plotting the trend over hours
ggplot(sightings_by_hour, aes(x = as.numeric(Hour), y = Sightings)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
  labs(title = "UFO Sightings Near Area 51 by Hour",
       x = "Hour of Day", y = "Number of Sightings") +
  theme_minimal()

```

### Combining Date and Time Trends


```{r warning=FALSE, message=FALSE}

# Grouping by hour and month and count sightings
sightings_by_time <- area51_sightings %>%
  group_by(Month, Hour) %>%
  summarise(Sightings = n())

# Plotting a heatmap
ggplot(sightings_by_time, aes(x = as.numeric(Month), y = as.numeric(Hour), fill = Sightings)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "UFO Sightings Near Area 51 by Month and Hour",
       x = "Month", y = "Hour of Day", fill = "Sightings") +
  theme_minimal()

```
### Adding Time Filters to Spatial Data


```{r}

# Filtering for July evenings (8 PM–10 PM)
july_evenings <- area51_sightings %>%
  filter(Month == "07" & Hour >= 20 & Hour <= 22)

# Plotting heat map for July evenings
ggplot(july_evenings, aes(x = longitude, y = latitude)) +
  geom_density2d() +
  stat_density2d(aes(fill = ..level..), geom = "polygon", alpha = 0.5) +
  scale_fill_gradient(low = "lightblue", high = "red") +
  labs(title = "UFO Sightings in July Evenings Near Area 51",
       x = "Longitude", y = "Latitude") +
  theme_minimal()

```
```{r}

coordinates <- data.frame(
  longitude = july_evenings$longitude,
  latitude = july_evenings$latitude,
  Region = july_evenings$Region,
  Locale = july_evenings$Locale
)

# Computing density
density_estimation <- kde2d(coordinates$longitude, coordinates$latitude, n = 100)

# Converting density grid into a data frame
density_data <- as.data.frame(expand.grid(
  longitude = density_estimation$x,
  latitude = density_estimation$y
))

# Adding density levels
density_data$density <- as.vector(density_estimation$z)

# Step 1: Filter for high-density regions (top 5% density values)
high_density_points <- density_data %>%
  filter(density >= quantile(density, 0.95))

# Step 2: Removing duplicates for latitude and longitude
high_density_unique <- high_density_points %>%
  distinct(longitude, latitude, .keep_all = TRUE)


```

### Ploting UFO Sightings Near Area 51 During July-Evenings

```{r}

# Perform clustering on latitude and longitude
coordinates <- july_evenings[, c("longitude", "latitude")]
dbscan_result <- dbscan(coordinates, eps = 0.01, minPts = 5)

# Add cluster labels to data
july_evenings$Cluster <- factor(dbscan_result$cluster)

# Plot clusters
ggplot(july_evenings, aes(x = longitude, y = latitude, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "UFO Sightings Clusters Near Area 51 During July-eveings.",
       x = "Longitude", y = "Latitude", color = "Cluster") +
  theme_minimal()
```
### UFO Sightings Clusters with Connections Near Area 51 During July-Evenings

```{r warning=FALSE, message=FALSE}

# Connect points by cluster (lines)
july_evenings_cluster_lines <- plot_ly(
  data = july_evenings,
  x = ~longitude,
  y = ~latitude,
  color = ~Cluster,
  type = 'scatter',
  mode = 'markers+lines',  # Add lines between points
  marker = list(size = 10, opacity = 0.8),
  text = ~paste(
    "Cluster:", Cluster, "<br>",
    "Longitude:", longitude, "<br>",
    "Latitude:", latitude, "<br>",
    "Locale:", Locale
  )
) %>%
  layout(
    title = "UFO Sightings Clusters with Connections Near Area 51 During July-Eveings",
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude"),
    hovermode = "closest"
  )

# Display the interactive plot
july_evenings_cluster_lines
```

## Performing Clustering at UFO Sightings Near Area 51

```{r}

# Performing clustering on latitude and longitude
coordinates <- area51_sightings[, c("longitude", "latitude")]
dbscan_result <- dbscan(coordinates, eps = 0.01, minPts = 5)

# Adding cluster labels to data
area51_sightings$Cluster <- factor(dbscan_result$cluster)

# Plotting clusters
ggplot(area51_sightings, aes(x = longitude, y = latitude, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "UFO Sightings Clusters Near Area 51",
       x = "Longitude", y = "Latitude", color = "Cluster") +
  theme_minimal()


```
## UFO Sightings Clusters with Connections Near Area 51 

```{r warning=FALSE, message=FALSE}


# Connecting points by cluster (lines)
plotly_cluster_lines <- plot_ly(
  data = area51_sightings,
  x = ~longitude,
  y = ~latitude,
  color = ~Cluster,
  type = 'scatter',
  mode = 'markers+lines',  
  marker = list(size = 10, opacity = 0.8),
  text = ~paste(
    "Cluster:", Cluster, "<br>",
    "Longitude:", longitude, "<br>",
    "Latitude:", latitude
  )
) %>%
  layout(
    title = "UFO Sightings Clusters with Connections Near Area 51",
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude"),
    hovermode = "closest"
  )

# Displaying the interactive plot
plotly_cluster_lines


```

